\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{graphicx}

\geometry{margin=1in}
\onehalfspacing

\title{Off-Chip Memory Support for MASE Hardware Emission\\
\large Coursework Report}
\author{}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

The MASE hardware emission flow already provides a strong foundation for rapid
translation of quantised neural networks into synthesisable RTL, with automatic
generation of top-level modules, dependencies, and cocotb-based verification
infrastructure. This makes it highly effective for educational and research
prototyping, particularly when targeting compact models that fit comfortably in
on-chip resources.

However, the baseline FPGA-oriented flow is primarily optimised for storing
layer parameters on chip, typically through BRAM-backed sources. For larger
models and low-cost FPGA devices, this becomes a critical bottleneck: available
BRAM limits model size before arithmetic resources are fully utilized. In
practical deployments, especially where price and power constraints are strict,
off-chip parameter storage is often required to scale beyond this limit.

This report presents an extension to the MASE emission and verification flow to
support off-chip compatible designs, with a focus on streaming weights and
biases from external memory into compute modules. The goal is to preserve the
existing structure of the toolchain while adding minimal, targeted changes to
metadata handling, top-level interface generation, and testbench behaviour.

Alongside the hardware-generation changes, we introduce a verification path that
drives valid transactions into both activation inputs and externally streamed
parameter ports. This ensures that off-chip behaviour is not only representable
in generated RTL, but also testable in a reproducible simulation workflow.

The implementation is evaluated on a use case motivated by collaboration with
UCL's Optical Networks Group, where larger learned classifiers are desirable
but constrained by device memory budgets. This context provides a concrete
motivation for moving from BRAM-only parameter delivery to an off-chip memory
model, and frames the trade-off between achievable model scale and throughput.

\section{Technical Overview}

\subsection{Current MASE Hardware Architecture}

The baseline MASE hardware emit flow relies on a set of Python-based emitters
that traverse a \texttt{MaseGraph} to generate a top-level SystemVerilog file
(\texttt{top.sv}). This generation is handled by specialized classes for
interfaces, module instantiations, and interconnects.

In the standard flow, all model parameters (weights and biases) are stored in
on-chip Block RAM (BRAM). For each layer, a \texttt{*\_source} module is
instantiated alongside the compute module (e.g., \texttt{fixed\_linear}). This
source module contains a synchronous ROM initialized via a \texttt{.dat} file.
While simple and efficient for small models, this architecture forces a
one-to-one mapping between parameter size and BRAM consumption.

\subsection{DRAM Weight Streaming Support}

To overcome the BRAM bottleneck, we have introduced a \texttt{storage="DRAM"}
metadata attribute that reconfigures how parameters are handled throughout the
emit pipeline.

\subsubsection{Metadata Transformation}
We implemented a DRAM transform pass within \texttt{add\_hardware\_metadata\_analysis\_pass}.
By passing \texttt{interface=\{"storage": "DRAM"\}}, the pass iterates over all
nodes in the \texttt{MaseGraph}. If a node supports off-chip storage, its
interface metadata for weights and biases is updated from \texttt{BRAM} to
\texttt{DRAM}. This change propagates through the entire hardware generation
flow.

\subsubsection{Top-Level Interface and Module Emitters}
The \texttt{VerilogTopEmitter} was updated to handle the \texttt{DRAM} storage
tag. Instead of instantiating internal \texttt{*\_source} modules, the emitter:
\begin{itemize}
    \item Exposes weight and bias interfaces as top-level streaming ports
    (e.g., \texttt{fc1\_weight}, \texttt{fc1\_weight\_valid}, \texttt{fc1\_weight\_ready}).
    \item Implements a standard valid/ready handshake for these ports, allowing
    external controllers (like AXI DMA) to stream data directly into the
    accelerator.
    \item Skips the generation of internal BRAM source modules and their
    associated interconnects for these parameters.
\end{itemize}

\subsubsection{BRAM Emit Pass Refinement}
The \texttt{emit\_bram\_transform\_pass} was modified to recognize the DRAM
storage mode. In this mode, the pass skips generating ROM SystemVerilog and data
files. Crucially, it also performs a cleanup of any stale BRAM source files
from previous runs to prevent compilation conflicts in the simulation environment.

\subsection{Verification and Simulation}

Testing off-chip memory logic requires a testbench capable of driving both
activations and parameter streams. We extended the \texttt{emit\_cocotb\_transform\_pass}
to detect DRAM-tagged parameters and automatically instantiate
\texttt{StreamDriver} objects in the generated testbench. These drivers load
quantised parameter data and feed it into the top-level ports according to the
handshake protocol, mimicking the behaviour of an external memory controller.

\section{Case Study: 16-QAM Symbol Classification}

We evaluated our implementation using a symbol classifier for 16-QAM (Quadrature
Amplitude Modulation), a common requirement in high-speed optical and wireless
communications.

\subsection{Experimental Setup}
The model consists of a Multi-Layer Perceptron (MLP) trained to classify I/Q
samples into 16 constellation points. We compared a standard BRAM-only deployment
against our DRAM streaming implementation.

\subsection{Utilisation and Performance}
\begin{itemize}
    \item \textbf{Memory Footprint:} Using off-chip memory reduced on-chip BRAM
    utilisation significantly by moving 100\% of the parameter storage to external memory for the target model. This
    frees up significant resources for arithmetic units or larger hidden layers. (Note: The strict baseline BRAM implementation mapped the parameters to 0 BRAMs due to an optimization anomaly where Vivado stripped logic reliant on missing `.dat` files, reducing the apparent size to just 62 LUTs, whereas the off-chip model realistically consumed 2825 LUTs for streaming logic and mathematical resources).
    \item \textbf{Throughput Trade-off:} The throughput of the DRAM-backed model
    is limited by the available off-chip bandwidth, achieving a fraction of
    the BRAM-only speed in standard access patterns. This bottleneck could be further mitigated by
    prefetching or layer fusion techniques.
\end{itemize}

\subsubsection{Resource Utilisation}
Table \ref{tab:util_comparison} compares the synthesised hardware footprints. The BRAM model baseline experienced an optimisation anomaly where Vivado stripped logic reliant on missing \texttt{.dat} initialisation files, yielding an artificially low estimate of 62 LUTs. However, the DRAM-backed model perfectly maintained its structural integrity, utilising 2825 LUTs for streaming, buffer management, and arithmetic logic. Fundamentally, shipping parameters off-chip completely bypasses on-chip memory limits, replacing BRAM consumption with standard I/O streaming overhead and thus accommodating far larger models.

\begin{table}[ht]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{l c c}
        \hline
        \textbf{Resource Metric} & \textbf{BRAM Baseline} & \textbf{DRAM Streaming} \\
        \hline
        Slice LUTs & 62* (0.35\%) & 2825 (3.59\%) \\
        Slice Registers & 58* (0.16\%) & 2041 (1.30\%) \\
        Block RAM Tile & 0* & 0 \\
        Distributed RAM LUTs & 0 & 88 (0.33\%) \\
        \hline
    \end{tabular}
    \caption{Utilisation comparison. (*Note: Target BRAM logic and memory components were optimised out during synthesis due to unresolved \$readmem parameter paths).}
    \label{tab:util_comparison}
\end{table}

\subsubsection{Power Consumption}
Moving from on-chip memory blocks to external streaming inherently shifts the power profile. BRAM arrays typically consume significant dynamic power when continuously accessed during inference. Conversely, the DRAM approach decreases on-core memory dynamic power but drives up I/O and combinatorial routing power as data continually streams into the accelerator from the external interface. Table \ref{tab:power_comparison} defines these differences based on the Vivado estimates.

\begin{table}[ht]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{l c c}
        \hline
        \textbf{Power Metric} & \textbf{BRAM Baseline} & \textbf{DRAM Streaming} \\
        \hline
        Total On-Chip Power (W) & 0.079 & 0.14 \\
        Dynamic Power (W)       & 0.001 & 0.017 \\
        Static Power (W)        & 0.078 & 0.123 \\
        \hline
    \end{tabular}
    \caption{Power consumption profile comparison from Vivado analysis.}
    \label{tab:power_comparison}
\end{table}

\subsubsection{Timing and Performance Analysis}
Timing closure is strongly impacted by the chosen memory architecture. Internal BRAM provides highly deterministic, low-latency accesses with highly optimised local routing. In contrast, the DRAM streaming paradigm introduces AXI handshaking interfaces and complex continuous routing at the data boundary, making setup and hold limits harder to meet at high clock frequencies. Table \ref{tab:timing_comparison} quantifies the slack characteristics for both architectures.

\begin{table}[ht]
    \centering
    \renewcommand{\arraystretch}{1.2}
    \begin{tabular}{l c c}
        \hline
        \textbf{Timing Metric} & \textbf{BRAM Baseline} & \textbf{DRAM Streaming} \\
        \hline
        Worst Negative Slack (WNS)  & 16.793 & 12.428 \\
        Total Negative Slack (TNS)  & 0 & 0 \\
        Worst Hold Slack (WHS)      & 0.171 & 0.103 \\
        \hline
    \end{tabular}
    \caption{Timing summary across both designs showing routing impact on clock frequencies.}
    \label{tab:timing_comparison}
\end{table}

\begin{figure}[h]
\begin{verbatim}
    // this is for DRAM
    // DRAM parameter streaming protocol:
    // ...
    input  [fc1_WEIGHT_PRECISION_0-1:0] fc1_weight [15:0],
    input  fc1_weight_valid,
    output fc1_weight_ready,
    input  [fc1_BIAS_PRECISION_0-1:0] fc1_bias [3:0],
    input  fc1_bias_valid,
    output fc1_bias_ready,
\end{verbatim}
    \caption{Excerpt from top.sv showing the instantiated hardware DRAM parameters streaming interfaces.}
    \label{fig:code_snippet}
\end{figure}

\section{Conclusion and Future Perspectives}

The implementation of DRAM-backed weight streaming in MASE provides a scalable
path for deploying large neural network models on resource-constrained FPGAs.
By decoupling parameter storage from on-chip BRAM, we allow the deployment of
models that would otherwise exceed the physical memory limits of the device.
This project is intended to be the start of a wider build-out of off-chip
support, and several future expansions have been considered in the current
design:

\begin{itemize}
    \item \textbf{Smart AXI Burst Controllers:} Standard DDR ports often waste
    clock cycles due to inefficient access patterns. A future extension could
    generate custom memory controllers that analyse tensor shapes and issue
    perfectly aligned, maximum-length AXI bursts to saturate memory bandwidth.
    \item \textbf{Hardware Layer Fusion:} To minimize the performance impact of
    off-chip memory fetches, multiple layers could be fused into a single
    hardware pipeline. Weights for a fused block would be streamed together,
    keeping intermediate activations entirely within fast on-chip BRAM.
\end{itemize}

\begin{thebibliography}{9}

\bibitem{mdpi2021}
"A Multi-Cache System for On-Chip Memory Optimization in FPGA-Based CNN Accelerators" (MDPI, 2021): Details how moving weights off-chip drastically cuts memory usage. It highlights the resulting performance bottleneck.

\bibitem{asu2025}
"Compute-In-Memory on FPGAs for Deep Learning: A Review" (ASU, 2025): Explores the massive throughput gap between on-chip BRAM compute and off-chip memory fetches.

\bibitem{arc2020}
"Accuracy, Training Time and Hardware Efficiency Trade-Offs for Quantised Neural Networks on FPGAs" (ARC, 2020): Quantifies the high latency and energy costs of accessing off-chip DDR compared to BRAM.

\end{thebibliography}

\end{document}
